\Large\textbf{Advection versus diffusion in brain ventricular transport}

### 2025-04-11 paper/main.tex
introduction

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{fame2020emergence}. Produced mostly by the choroid
plexus and partly by brain interstitial fluid,
CSF is a clear liquid that flows through the ventricular system and subarachnoid
space~\cite{Brodal2016, Damkier2013}. By transporting growth factors, nutrients, neurotransmitters
and metabolites, as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{Ringers2020Role, fame2020emergence, del2010ependymal}.
Disruption of the CSF circulation is detrimental to brain health and disturbs
cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment~\cite{duy2022rethinking, kahle2024paediatric, wallmeier2022role},
but for which there is little agreement on optimal treatments~\cite{flannery2014pediatric}.

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure
gradients generated by the cardiac cycle, respiration, neural
activity, and body movements~\cite{Brodal2016, del2010ependymal,
  Ringers2020Role, Vinje2019RespiratoryMeasurements,
  Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
  macaulay2022cerebrospinal, mestre2018flow}.  Cilia are hair-like
structures found in a vast diversity of biological organisms and in
several human organs, including the brain~\cite{DGama2025MotileBrain,
  Fulford1986Muco-ciliaryLung, Pacherres2022CiliaryProduction,
  Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella,
  Ringers2020Role, spassky2013motile, Pellicciotta2020Cilia,
  mitchell2007evolution, Tsukita2012CoordinatedFeet,
  BLAKE1974MechanicsMotion,
  Reiten2017Motile-Cilia-MediatedComputations,
  Thouvenin2020OriginCanal}.  The ventricular surface of the brain,
known as the ependyma, is lined by motile cilia that beat in a
coordinated pattern at frequencies of 10--40 Hz to mediate fluid
flow~\cite{mitchell2007evolution, Ringers2020Role, spassky2013motile,
  Ringers2023NovelEpithelia, roth2025structure}.  Given their
microscopic size, ranging from 5 to 15 microns, cilia contribute
primarily to flow near the ependyma in the human
brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
  Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast,
cardiac, respiratory and delta waves at lower frequencies induce
pulsatile flow at the larger scales in the brain ventricular system,
subarachnoid space, and subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
  Causemann2025, fultz2019coupled, mestre2018flow,
  daversin2020mechanisms}.

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.
To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental approaches face technical challenges: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they did not study the role of other contributions to CSF
flow and the accompanying transport, such as pressure
gradients associated with respiration and cardiac pulsations. Moreover, while MRI technology
can measure CSF motion in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. 
These technical limitations of studying cilia-mediated flow and transport make 
mathematical modeling and computational investigations an attractive complement to experimental
studies. Previous simulation studies have focused on modeling the interaction
between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}, but also 
other applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella}, and
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.
However, we still lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport within the brain.
This is of particular importance since disorders such as hydrocephalus are observed
in animal models and humans with deficient motile cilia~\cite{Afzelius2004CiliaRelatedDiseases,
Olstad2019CiliaryDevelopment, Mitchison2017Motile, Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.

>>>>

Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{fame2020emergence}. Produced predominantly by the choroid
plexus, CSF is a clear liquid that flows through the ventricular system and subarachnoid
space~\cite{Brodal2016, Damkier2013, Wichmann2022Overview}.
By transporting growth factors, nutrients, neurotransmitters
and metabolites, as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{Ringers2020Role, fame2020emergence, del2010ependymal}.
Disruption of CSF circulation is detrimental to brain health and disturbs
cognitive and motor functions~\cite{Johanson2008},
and accumulation of CSF leads to hydrocephalus,
a disorder often requiring emergency surgical treatment.
\cite{duy2022rethinking, kahle2024paediatric, wallmeier2022role,
flannery2014pediatric}

The flow of CSF in the ventricular system is driven by multiple
physiological processes, including the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure
gradients generated by the cardiac cycle, respiration, neural
activity, and body movements~\cite{Brodal2016, del2010ependymal,
Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow, Williams2023NeuralActivity}. 
Cilia are hair-like structures found in a vast diversity of
biological organisms and in
several human organs, including the brain~\cite{DGama2025MotileBrain,
Fulford1986Muco-ciliaryLung, Pacherres2022CiliaryProduction,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella,
Ringers2020Role, spassky2013motile, Pellicciotta2020Cilia,
mitchell2007evolution, Tsukita2012CoordinatedFeet,
BLAKE1974MechanicsMotion,
Reiten2017Motile-Cilia-MediatedComputations,
Thouvenin2020OriginCanal}. Notably, motile cilia
are found on the ventricular surface of the brain,
known as the ependyma, and beat in a
coordinated pattern at frequencies of 10--40 Hz to mediate fluid
flow~\cite{mitchell2007evolution, Ringers2020Role, spassky2013motile,
Ringers2023NovelEpithelia, roth2025structure}. Given their
microscopic size, ranging from 5 to 15 microns, cilia contribute
primarily to flow near the ependyma in the
brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast,
cardiac, respiratory and delta waves at lower frequencies induce
pulsatile flow at larger scales in the brain ventricular system,
subarachnoid space, and perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow,
daversin2020mechanisms}.

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.
To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental approaches face technical challenges: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other slower fluid dynamic processes that span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in the rodent and human brain.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they did not study the role of other contributions to CSF
flow and the accompanying transport, such as pressure
gradients associated with respiration and cardiac pulsations. Moreover, while MRI technology
can measure CSF motion in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. 
These technical limitations of studying cilia-mediated flow and transport make 
mathematical modeling and computational investigations an attractive complement to experimental
studies. Previous simulation studies have focused on modeling the interaction
between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal},
and other applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella}, and
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.
However, we still lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport within the brain.
This is of particular importance since disorders such as hydrocephalus are observed
in animal models and humans with deficient motile cilia~\cite{Afzelius2004CiliaRelatedDiseases,
Olstad2019CiliaryDevelopment, Mitchison2017Motile, Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.
